---
# Get YAML keywords from myYAML_ref.Rmd
title: "Motor Trend: mpg regression"
author: "bdanalytics"
#output: html_document
---
#### Date: `r format(Sys.time(), "(%a) %b %d, %Y")`

Data: Motor Trend Car Road Tests "mtcars {datasets}"  
Source: Henderson and Velleman (1981), Building multiple regression models interactively. Biometrics, 37, 391–411.  
Time period: 1973–74 models

### Synopsis:

#### Is an automatic or manual transmission better for MPG:
Manual transmission is better for miles per gallon versus automatic transmission.  
Average mpg for automatic transmission is `17` vs. `23` for manual transmission.  

#### Quantify the MPG difference between automatic and manual transmissions:

#### Potential next steps include:

```{r set_global_options}
rm(list=ls())
set.seed(12345)
source("~/Dropbox/datascience/R/mydsutils.R")
source("~/Dropbox/datascience/R/myplot.R")
# Gather all package requirements here
#suppressPackageStartupMessages(require())
```

```{r set_global_options_wd, echo=FALSE}
setwd("~/Documents/Work/Courses/Coursera/jhu-datascience/G-Regression-Models/Project/MotorTrend")
```

### Step 01: import data
```{r import_data, cache=TRUE}
data(mtcars)
cars_df <- mtcars
myprint_df(cars_df)
```

#### Step 02.1: inspect data
Automtatic Transmission feature (am) is numeric. Let's make it a factor for analytics convenience.
```{r inspect_steps_1, cache=TRUE}
cars_df$am_fctr <- factor(cars_df$am, labels=c("automatic", "manual"))
print(str(cars_df))

#[, 1]  mpg	    Miles/(US) gallon
#[, 2]	cyl	    Number of cylinders
#[, 3]	disp	Displacement (cu.in.)
#[, 4]	hp	    Gross horsepower
#[, 5]	drat	Rear axle ratio
#[, 6]	wt	    Weight (lb/1000)
#[, 7]	qsec	1/4 mile time
#[, 8]	vs	    V/S
#[, 9]	am	    numeric:    Transmission (0 = automatic, 1 = manual)
#[,12]  am_fctr	factor:     Transmission (automatic, manual)
#[,10]	gear	Number of forward gears
#[,11]	carb	Number of carburetors

print(summary(cars_df))
#pairs(subset(entity_df, select=-c(col_symbol)))

myplot_violin <- function(df, ycol_names, xcol_name=NULL, facet_spec=NULL) {
    if ((length(ycol_names) > 1) & (!missing(xcol_name)))
        stop("Multiple feats not implemented with x variable.", 
             "\n  Consider using facet parameter instead.")
    
    if (!missing(xcol_name) & !is.factor(df[, xcol_name])) {
        xcol_name_par <- xcol_name
        xcol_name <- paste(xcol_name_par, "fctr", sep="_")
        warning("xcol_name:", xcol_name_par, " is not a factor; creating ", xcol_name)
        df[, xcol_name] <- as.factor(df[, xcol_name_par])
    }
    
    if (length(ycol_names) == 1) {
        if (missing(xcol_name)) {
            medians_df <- summaryBy(as.formula(paste0(ycol_names, " ~ factor(0)")), df, 
                                    FUN=c(median), na.rm=TRUE)
            p <- ggplot(df, aes_string(x=factor(0), y=ycol_names))
            p <- p + xlab(" ")            
        } else {
            medians_df <- summaryBy(as.formula(paste0(ycol_names, " ~ ", xcol_name)), df, 
                                    FUN=c(median), na.rm=TRUE)
            p <- ggplot(df, aes_string(x=xcol_name, y=ycol_names))            
        }
    } else {
        require(reshape)
        mltd_df <- melt(df,, measure.vars=ycol_names)
        require(doBy)        
        medians_df <- summaryBy(value ~ variable , mltd_df, FUN=c(median), na.rm=TRUE)
        p <- ggplot(mltd_df, aes(x=variable, y=value))
        p <- p + xlab(" ")
    }
    
    if (!is.null(facet_spec)) {
        stop("facets not supported yet")
        require(doBy)
        sum_df <- summaryBy(steps ~ . , df, FUN=c(median))
    } else {
    }
    
    p <- p + geom_violin(fill="grey80", color="blue") + 
             stat_summary(fun.y=mean, pch=22, geom='point', color='red') +
             scale_y_continuous(labels=myformat_number)
    
    if (length(ycol_names) == 1) {
        aes_str <- paste0("linetype=\"dashed\", yintercept=as.numeric(", ycol_names, ")")        
        aes_mapping <- eval(parse(text = paste("aes(", aes_str, ")")))
#         p <- p + geom_hline(data=medians_df, 
#                             mapping=aes_mapping, show_guide=TRUE
#                             , color="black", size=1)
#         p <- p + scale_linetype_identity(guide="legend", name="Stats", 
#                                          labels=rownames(medians_df))
        
        aes_str <- paste0("y=", ycol_names, ".median * 1.05", 
                          ", label=myformat_number(round(", ycol_names, ".median))")
        aes_mapping <- eval(parse(text = paste("aes(", aes_str, ")")))
        p <- p + geom_text(data=medians_df, 
                           mapping=aes_mapping
                           , color="NavyBlue", size=3.5)
    } else {
        #print(medians_df)
        p <- p + geom_text(data=medians_df,
                           mapping=aes_string(x="variable", 
                                              y="value.median * 1.05",
                                              label="myformat_number(round(value.median))")
                           , color="NavyBlue", size=3.5)
        #print("median text layer applied")
    }
    
    return(p)
}
print(myplot_violin(cars_df, "mpg", "am_fctr"))

# Create new features that help diagnostics
```

#### Is an automatic or manual transmission better for MPG:
Null Hypothesis ($\sf{H_{0}}$): mpg is not impacted by am_fctr.  
The variance by am_fctr appears to be independent. 
```{r q1, cache=TRUE}
print(t.test(subset(cars_df, am_fctr == "automatic")$mpg, 
             subset(cars_df, am_fctr == "manual")$mpg, 
             var.equal=FALSE)$conf)
```
We reject the null hypothesis i.e. we have evidence to conclude that am_fctr impacts mpg (95% confidence). Manual transmission is better for miles per gallon versus automatic transmission.

```{r print_sessionInfo, echo=FALSE}
sessionInfo()
```