---
title: 'Motor Trend: mpg regression'
author: "bdanalytics"
output: pdf_document
---
#### Date: `r format(Sys.time(), "(%a) %b %d, %Y")`

Data: Motor Trend Car Road Tests "mtcars {datasets}"  
Source: Henderson and Velleman (1981), Building multiple regression models interactively. Biometrics, 37, 391–411.  
Time period: 1973–74 models

```{r set_global_options, echo=FALSE}
rm(list=ls())
set.seed(12345)
source("~/Dropbox/datascience/R/mydsutils.R")
source("~/Dropbox/datascience/R/myplot.R")
# Gather all package requirements here
suppressPackageStartupMessages(require(reshape2))
suppressPackageStartupMessages(require(plyr))
```

```{r set_global_options_wd, echo=FALSE}
setwd("~/Documents/Work/Courses/Coursera/jhu-datascience/G-Regression-Models/Project/MotorTrend")
```
### Synopsis:

#### Is an automatic or manual transmission better for MPG:
Manual transmission is better for miles per gallon versus automatic transmission.  
Average mpg for automatic transmission is `17` vs. `23` for manual transmission.  

#### Quantify the MPG difference between automatic and manual transmissions:
The univariate model yields `mpg ~ 17.15 + 7.25 * manual` while explaining only `33.8%` of the mpg variation.

The proposed multivariate model yields `mpg ~ 46.30 - 9.08 * wt` for cars with manual transmission and `mpg ~ 31.42 - 3.79 * wt` for cars with automatic transmission where `wt` is weight (lb/1000) of the car. This model explains `81.5%` of the mpg variation at a `99%` confidence level. The most influential data point turned out to be Maserati Bora which would increase the predicted mpg by `0.63` for cars with manual transmission without it. The most influential data point for automatic transmission was Lincoln Continental.

```{r synopsis_plot_model_wt_i, echo=FALSE, fig.height=5}
data(mtcars)
cars_df <- mtcars
cars_df$am_fctr <- factor(cars_df$am, labels=c("automatic", "manual"))
cars_df$id <- rownames(cars_df)
cars_df <- mutate(cars_df, 
                  id_outlier=ifelse(((id == "Maserati Bora") | 
                                     (id == "Lincoln Continental")), id, ""))
print(myplot_scatter(cars_df, "wt", "mpg", colorcol_name="am_fctr") + 
      #xlim(0, 6) +    
      scale_x_continuous(limits=c(1.5, 5.5), breaks=seq(1.5, 5.5, by=0.5)) +       
      geom_text(aes(label=id_outlier), size=3)      
     )
```

The model for variation from mean weight (wt) did not pass the statistical significance tests. Additional features were not statistically significant and/or explain additional mpg variation. The proposed model contains minor negative correlation of residuals with predicted values & residual heteroskedacity.

#### Potential next steps include:
1. Compress report to 5 pages; Knit PDF keeps crashing on my computer... extremely tedious to optimize length
2. Test other regression techniques (e.g. additive models in glm) to better quantify the relationship.

### Appendix:
#### Import data & setup analytics:
```{r import_data, cache=TRUE, echo=FALSE, results='hide'}
data(mtcars)
cars_df <- mtcars
myprint_df(cars_df)
```

Automatic Transmission feature (am), number of cylinders (cyl), V/S (vs), number of forward gears (gear) & number of carburetors (carb) are numeric. Let's make them factors for analytics convenience.

```{r inspect_data_1, cache=TRUE, echo=FALSE, results='hide', fig.height=5}
cars_df$am_fctr <- factor(cars_df$am, labels=c("automatic", "manual"))
cars_df$cyl_fctr <- factor(cars_df$cyl)
cars_df$vs_fctr <- factor(cars_df$vs, labels=c("V", "S"))
cars_df$gear_fctr <- factor(cars_df$gear)
cars_df$carb_fctr <- factor(cars_df$carb)
cars_df$id <- rownames(cars_df)
cars_df <- mutate(cars_df, 
                  id_outlier=ifelse(((id == "Maserati Bora") | 
                                     (id == "Lincoln Continental")), id, ""))
#print(str(cars_df))

#[, 1]  mpg	        numeric:    Miles/(US) gallon
#[, 2]	cyl	        numeric:    Number of cylinders
#[,13]  cyl_fctr	factor:     Number of cylinders
#[, 3]	disp	    numeric:    Displacement (cu.in.)
#[, 4]	hp	        numeric:    Gross horsepower
#[, 5]	drat	    numeric:    Rear axle ratio
#[, 6]	wt	        numeric:    Weight (lb/1000)
#[, 7]	qsec	    numeric:    1/4 mile time
#[, 8]	vs	        numeric:    V/S
#[,14]  vs_fctr     factor:     V/S (0 = V, 1 = S) ???
#[, 9]	am	        numeric:    Transmission (0 = automatic, 1 = manual)
#[,12]  am_fctr	    factor:     Transmission (automatic, manual)
#[,10]	gear	    numeric:    Number of forward gears
#[,15]  gear_fctr	factor:     Number of forward gears
#[,11]	carb	    numeric:    Number of carburetors
#[,16]  carb_fctr	factor:     Number of carburetors

#[,17]  id          string:     Car make & model

print(summary(cars_df))

feats_num <- c("disp", "hp", "drat", "wt", "qsec")
feats_fctr <- c("am_fctr", "cyl_fctr", "vs_fctr", "gear_fctr", "carb_fctr")
```
```{r inspect_data_2, cache=TRUE, echo=FALSE}
print(myplot_violin(cars_df, "mpg", "am_fctr") + 
      geom_text(aes(label=id_outlier), size=3)
     )
```

#### Is an automatic or manual transmission better for MPG:
Null Hypothesis ($\sf{H_{0}}$): mpg is not impacted by am_fctr.  
The variance by am_fctr appears to be independent. 
```{r q1, cache=TRUE}
print(t.test(subset(cars_df, am_fctr == "automatic")$mpg, 
             subset(cars_df, am_fctr == "manual")$mpg, 
             var.equal=FALSE)$conf)
```
We reject the null hypothesis i.e. we have evidence to conclude that am_fctr impacts mpg (95% confidence). Manual transmission is better for miles per gallon versus automatic transmission.

#### Quantify the MPG difference between automatic and manual transmissions:
Let's try the univariate model to establish a benchmark against which we can evaluate more complex models, if necessary
```{r q2_test_model_1, cache=TRUE}
mpg_fit <- lm(mpg ~ am_fctr, data=cars_df)
print(summary(mpg_fit))
```
This univariate model is statistically significant and explains ``r format(summary(mpg_fit)$adj.r.squared * 100.0, digits=3, nsmall=1)``% of the mpg variation. 

Manual Transmission provides ``r format(mpg_fit$coeff[2], digits=3)`` additional miles per gallon compared to automatic transmission. This result is highly significant - 95% confidence interval is `[`r format(confint(mpg_fit)[2, ], digits=3)`]`.

Let's inspect the residuals for any bias 
```{r q2_inspect_model_1, cache=TRUE, echo=FALSE, fig.height=5}
build_fit_df <- function(fit) {
    df <- data.frame(mpg.predict=predict(fit, cars_df), 
                     mpg.residuals=fit$residuals,
                     mpg=cars_df$mpg, am_fctr=cars_df$am_fctr, 
                     id=cars_df$id, id_outlier=cars_df$id_outlier)
    df <- cbind(df, cars_df[, feats_fctr[feats_fctr != "am_fctr"]])
    df <- cbind(df, cars_df[, feats_num[feats_num != "mpg"]])
}
mpg_fit_df <- build_fit_df(mpg_fit)

print(myplot_violin(mpg_fit_df, "mpg.residuals", "mpg.predict") + 
      geom_text(aes(label=id_outlier), size=3)        
      )
```  

The residuals of the regression appear NOT biased with mean close to `0`. However, there is some residual heteroskedacity (median residuals for manual transmission is -2). Let's examine if the residuals are impacted by features not in this model. For that, let's first inspect the correlations between mpg & each of the features and build labels that are used later in the residual plots

```{r q2_compute_cor, cache=TRUE, echo=FALSE}
mpg_cor_df <- data.frame(mpg.cor=cor(cars_df[, c(gsub("_fctr", "", feats_fctr), 
                                                 feats_num)], cars_df[, "mpg"]))
mpg_cor_df$mpg.cor.abs <- abs(mpg_cor_df$mpg.cor)
mpg_cor_df <- orderBy(~ -mpg.cor.abs, data=mpg_cor_df)
mpg_cor_df$feature <- rownames(mpg_cor_df)

mpg_cor_df <- mutate(mpg_cor_df, 
                     variable=ifelse(feature %in% feats_num, feature, paste0(feature, "_fctr")),
                     label=paste0(ifelse(feature %in% feats_num, feature, paste0(feature, "_fctr")), ".mpg.cor=", ifelse(mpg.cor > 0, '+', ''),
                                  format(mpg.cor, nsmall=3, digits=3)))
myprint_df(mpg_cor_df)
```  

```{r q2_inspect_model_1_residuals, cache=TRUE, echo=FALSE, fig.height=4}

build_fit_plot_df <- function(fit_df) {
    fit_plot_df <- melt(fit_df[, c("id", "id_outlier", "mpg", "mpg.predict", "mpg.residuals", 
                                           feats_fctr, feats_num)], 
                            id.vars=c("id", "id_outlier", "mpg", "mpg.predict", "mpg.residuals", 
                                      "am_fctr"))
    fit_plot_df <- orderBy(~ id + variable, fit_plot_df)
    #tail(fit_df)
    #tail(fit_plot_df, 10)
    #intersect(names(fit_plot_df), names(mpg_cor_df[, c("variable", "label")]))
    fit_plot_df <- merge(fit_plot_df, mpg_cor_df[, c("variable", "label")],
                                 all.x=TRUE)
    fit_plot_df <- orderBy(~ id + variable, fit_plot_df)
    return(fit_plot_df)
}
mpg_fit_plot_df <- build_fit_plot_df(mpg_fit_df)

print(myplot_scatter(mpg_fit_plot_df, "mpg.predict", "mpg.residuals", 
                     colorcol_name="am_fctr", group=1) + 
      facet_wrap(~ label) + 
      geom_text(aes(label=id_outlier), size=3)  
     )
```

None of the other variables seem to explain any more mpg variation at first glance. Let's check the actual distribution of mpg vs. am_fctr.

```{r q2_inspect_data_1, cache=TRUE, echo=FALSE}
cars_plot_df <- melt(cars_df[, c("id", "id_outlier", "mpg", feats_fctr, feats_num)], 
                     id.vars=c("id", "id_outlier","mpg", "am_fctr"))
cars_plot_df <- orderBy(~ id + variable, cars_plot_df)
#tail(cars_df)
#tail(cars_plot_df)
#intersect(names(cars_plot_df), names(mpg_cor_df[, c("variable", "label", "mpg.cor.abs")]))
cars_plot_df <- merge(cars_plot_df, mpg_cor_df[, c("variable", "label", 
                                                   "mpg.cor.abs")],
                      all.x=TRUE)
cars_plot_df <- orderBy(~ id + variable, cars_plot_df)
cars_plot_df$label_fctr <- factor(cars_plot_df$label, levels=mpg_cor_df$label)
print(myplot_scatter(cars_plot_df, "value", "mpg", colorcol_name="am_fctr", group=1) + 
      facet_wrap(~ label_fctr, scales="free_x") + 
      geom_text(aes(label=id_outlier), size=3)  
     )
```

Clearly, other features can provide additional insights into the relationship between mpg & transmission. Let's add features to the simple model (mpg_fit) in order of the features correlated with mpg (highest tested first).

```{r q2_test_model_wt, cache=TRUE, echo=FALSE}
mpg_wt_fit <- update(mpg_fit, mpg ~ am_fctr + wt, data=cars_df)
mpg_wt_i_fit <- update(mpg_fit, mpg ~ am_fctr * wt, data=cars_df)
print(anova(mpg_fit, mpg_wt_fit, mpg_wt_i_fit))
print(summary(mpg_wt_i_fit))
```
The interaction model of am_fctr & wt is statistically significant (`99%` confidence) and all the model coefficients are significant (`99%` confidence). None of the other models tested piecewise for each additional feature to this model in a similar fashion crossed these thresholds. Adj-Rsq is ``r format(summary(mpg_wt_i_fit)$adj.r.squared, nsmall=3, digits=3)``. Let's inspect the residuals for this model. 

```{r q2_inspect_model_wt_i_1, echo=FALSE, fig.height=4}
mpg_wt_i_fit_df <- build_fit_df(mpg_wt_i_fit)

extract_fit_formula <- function(fit) {
    str <- strsplit(toString(summary(fit)$call$formula), ',')
    tmp_str <- str[[1]]
    str <- paste0(tmp_str[2], " ", tmp_str[1], tmp_str[3:length(tmp_str)])
}

violin_plot_df <- rbind(cbind(mpg_fit_df, 
                             data.frame(fit_str=rep(extract_fit_formula(mpg_fit), 
                                                    nrow(mpg_fit_df)))),
                        cbind(mpg_wt_i_fit_df, 
                              data.frame(fit_str=rep(extract_fit_formula(mpg_wt_i_fit), 
                                                     nrow(mpg_wt_i_fit_df)))))
print(ggplot(violin_plot_df, aes(x=am_fctr, y=mpg.residuals)) + 
      geom_violin(aes(fill=fit_str), color="black") + 
      facet_wrap(~ fit_str) + 
      stat_summary(fun.y=mean, pch=22, geom='point', color='red') + 
      geom_smooth(aes(group=1), method="lm") + 
      theme(legend.position="none") + 
      geom_text(aes(label=id_outlier), size=3)  
     )
```

The mean of residuals is `0` for both transmission types and the heteroskedacity of the residuals is reduced significantly. 

```{r q2_inspect_model_wt_i_2, echo=FALSE, fig.height=4}
# names(violin_plot_df)
print(ggplot(violin_plot_df, aes(x=mpg.predict, y=mpg.residuals)) + 
      geom_violin(aes(fill=fit_str), color="blue") + 
      facet_wrap(~ fit_str) + 
      stat_summary(fun.y=mean, pch=22, geom='point', color='red') + 
      geom_smooth(aes(group=1), method="rlm") + 
      theme(legend.position="none") + 
      geom_text(aes(label=id_outlier), size=3)  
      )
```

There seems to be negatively correlated residuals in the interactive wt model, although that doesn't show up when method="lm" in geom_smooth() 

```{r q2_inspect_model_wt_i_3, echo=FALSE}
# par(mfcol=c(1, 2))
# plot(mpg_fit, which=1)
# plot(mpg_wt_i_fit, which=1)

mpg_wt_i_fit_plot_df <- build_fit_plot_df(mpg_wt_i_fit_df)
scatter_plot_df <- rbind(cbind(mpg_fit_plot_df, 
                             data.frame(fit_str=rep(extract_fit_formula(mpg_fit), 
                                                    nrow(mpg_fit_df)))),
                         cbind(mpg_wt_i_fit_plot_df, 
                              data.frame(fit_str=rep(extract_fit_formula(mpg_wt_i_fit), 
                                                     nrow(mpg_wt_i_fit_df)))))
print(myplot_scatter(scatter_plot_df, "mpg.predict", "mpg.residuals", 
                     colorcol_name="am_fctr", group=1) + 
                     facet_grid(label ~ fit_str))

#plot(mpg_wt_i_fit, which=1)
plot(mpg_wt_i_fit, which=2)
print(shapiro.test(mpg_wt_i_fit_plot_df$mpg.residuals))
#plot(mpg_wt_i_fit, which=3)
#plot(mpg_wt_i_fit, which=4)
#plot(mpg_wt_i_fit, which=5)
hatvals_mpg_wt_i_fit <- hatvalues(mpg_wt_i_fit)
max_hatval_car <- which(hatvals_mpg_wt_i_fit == max(hatvals_mpg_wt_i_fit))
print("Max hat value:")
print(cars_df[max_hatval_car, ])
print(format(dfbeta(mpg_wt_i_fit)[max_hatval_car, ], digits=2, scientific=FALSE))

#plot(mpg_wt_i_fit, which=6)
```

The Maserati Bora is the most influential point and without it manual transmission would be ``r format((dfbeta(mpg_wt_i_fit)[max_hatval_car, ]["am_fctrmanual"]), digits=2)`` mpg (lower) - [am_fctrmanual] and ``r format((dfbeta(mpg_wt_i_fit)[max_hatval_car, ]["am_fctrmanual:wt"] * 3.57), digits=2)`` mpg (higher) [am_fctrmanual:wt].

```{r q2_print_model_wt_i, echo=FALSE}
build_fit_formula <- function(fit) {
    str <- strsplit(toString(summary(fit)$call$formula), ',')
    tmp_str <- str[[1]]
#     regressors <- strsplit(tmp_str[3], split=' ')[[1]]
#     regressors <- regressors[regressors != '+']
    coeff <- summary(fit)$coefficients
    coeff_str <- ""
    for (row in seq(1:nrow(coeff)))
    #for (row in seq(1:2))        
        coeff_str <- paste0(coeff_str, ifelse((row == 1), "", " + "), 
                            format(coeff[row, 1], digits=4), 
                            ifelse((row == 1), "", " * "), 
                            ifelse((row == 1), "", rownames(coeff)[row]))                            
    #print(coeff_str)
    str <- paste0(tmp_str[2], " ", tmp_str[1], " ", coeff_str)
    return(str)
}
display_fit_frmla <- build_fit_formula(mpg_wt_i_fit)
print(sprintf("Proposed model: %s", display_fit_frmla))
print(confint(mpg_wt_i_fit))
```

The proposed model is ``r display_fit_frmla`` where `am_fctrmanual` is 1 for manual transmission [0 for automatic], `wt` is weight (lb/1000) and `am_fctrmanual:wt` is wt for manual transmission [0 for automatic]. None of the coefficients change sign in the 95% confidence interval.

```{r q2_plot_model_wt_i, echo=FALSE, results='hide'}
#model_coeffs <- summary(mpg_wt_i_fit)$coefficients[, 1]
# print(myplot_scatter(cars_df, "wt", "mpg", colorcol_name="am_fctr"))
# print(myplot_scatter(cars_df, "wt", "mpg", colorcol_name="am_fctr") + 
#       geom_abline(intercept=(model_coeffs[1]), 
#                   slope=(model_coeffs[3]), color="red") +
#       geom_abline(intercept=(model_coeffs[1] + model_coeffs[2]), 
#                   slope=(model_coeffs[3] + model_coeffs[4]), color="blue")  
#      )
```

```{r print_sessionInfo, echo=FALSE}
sessionInfo()
```