---
# Get YAML keywords from myYAML_ref.Rmd
title: "Motor Trend: mpg regression"
author: "bdanalytics"
#output: html_document
---
#### Date: `r format(Sys.time(), "(%a) %b %d, %Y")`

Data: Motor Trend Car Road Tests "mtcars {datasets}"  
Source: Henderson and Velleman (1981), Building multiple regression models interactively. Biometrics, 37, 391–411.  
Time period: 1973–74 models

### Synopsis:

#### Is an automatic or manual transmission better for MPG:
Manual transmission is better for miles per gallon versus automatic transmission.  
Average mpg for automatic transmission is `17` vs. `23` for manual transmission.  

#### Quantify the MPG difference between automatic and manual transmissions:
Manual Transmission provides `7.2449` additional miles per gallon compared to automatic transmission. This result is highly significant - 95% confidence interval is `[3.6415, 10.8484]`

#### Potential next steps include:
0. Inspect residuals by features not used in model
1. Adjust model for residual heteroskedacity
2. Test other models to better quantify the relationship (confounding factors)

```{r set_global_options}
rm(list=ls())
set.seed(12345)
source("~/Dropbox/datascience/R/mydsutils.R")
source("~/Dropbox/datascience/R/myplot.R")
# Gather all package requirements here
#suppressPackageStartupMessages(require())
```

```{r set_global_options_wd, echo=FALSE}
setwd("~/Documents/Work/Courses/Coursera/jhu-datascience/G-Regression-Models/Project/MotorTrend")
```

### Step 01: import data
```{r import_data, cache=TRUE}
data(mtcars)
cars_df <- mtcars
myprint_df(cars_df)
```

#### Step 02.1: inspect data
Automtatic Transmission feature (am), number of cylinders (cyl), V/S (vs), number of forward gears (gear) & number of carburetors (carb) are numeric. Let's make them factors for analytics convenience.
```{r inspect_steps_1, cache=TRUE}
cars_df$am_fctr <- factor(cars_df$am, labels=c("automatic", "manual"))
cars_df$cyl_fctr <- factor(cars_df$cyl)
cars_df$vs_fctr <- factor(cars_df$vs, labels=c("V", "S"))
cars_df$gear_fctr <- factor(cars_df$gear)
cars_df$carb_fctr <- factor(cars_df$carb)
print(str(cars_df))

#[, 1]  mpg	        numeric:    Miles/(US) gallon
#[, 2]	cyl	        numeric:    Number of cylinders
#[,13]  cyl_fctr	factor:     Number of cylinders
#[, 3]	disp	    numeric:    Displacement (cu.in.)
#[, 4]	hp	        numeric:    Gross horsepower
#[, 5]	drat	    numeric:    Rear axle ratio
#[, 6]	wt	        numeric:    Weight (lb/1000)
#[, 7]	qsec	    numeric:    1/4 mile time
#[, 8]	vs	        numeric:    V/S
#[,14]  vs_fctr     factor:     V/S (0 = V, 1 = S) ???
#[, 9]	am	        numeric:    Transmission (0 = automatic, 1 = manual)
#[,12]  am_fctr	    factor:     Transmission (automatic, manual)
#[,10]	gear	    numeric:    Number of forward gears
#[,15]  gear_fctr	factor:     Number of forward gears
#[,11]	carb	    numeric:    Number of carburetors
#[,16]  carb_fctr	factor:     Number of carburetors

print(summary(cars_df))
#pairs(subset(entity_df, select=-c(col_symbol)))

print(myplot_violin(cars_df, "mpg", "am_fctr"))
feats_num <- c("disp", "hp", "drat", "wt", "qsec")
feats_fctr <- c("am_fctr", "cyl_fctr", "vs_fctr", "gear_fctr", "carb_fctr")

# Create new features that help diagnostics
```

#### Is an automatic or manual transmission better for MPG:
Null Hypothesis ($\sf{H_{0}}$): mpg is not impacted by am_fctr.  
The variance by am_fctr appears to be independent. 
```{r q1, cache=TRUE}
print(t.test(subset(cars_df, am_fctr == "automatic")$mpg, 
             subset(cars_df, am_fctr == "manual")$mpg, 
             var.equal=FALSE)$conf)
```
We reject the null hypothesis i.e. we have evidence to conclude that am_fctr impacts mpg (95% confidence). Manual transmission is better for miles per gallon versus automatic transmission.

#### Quantify the MPG difference between automatic and manual transmissions:

Let's try the univariate model to establish a benchmark against which we can evaluate more complex models, if necessary
```{r q2_1, cache=TRUE}
mpg_fit <- lm(mpg ~ am_fctr, data=cars_df)
print(summary(mpg_fit))
```
Manual Transmission provides ``r mpg_fit$coeff[2]`` additional miles per gallon compared to automatic transmission. This result is highly significant - 95% confidence interval is `[`r confint(mpg_fit)[2, ]`]`. About ``r round(summary(mpg_fit)$r.squared * 100, 2)``% of the mpg variance is attributable to transmission.

Let's inspect the residuals for any bias 
```{r q2_2, cache=TRUE}
#print(myplot_scatter(cars_df, "am_fctr", "mpg", colorcol_name="am_fctr"))

mpg_residuals_df <- data.frame(mpg_fit$residuals, cars_df$mpg, cars_df$am_fctr)
#print(summaryBy(mpg_fit.residuals ~ cars_df.am_fctr, data=mpg_residuals_df, FUN=mean))
print(myplot_violin(mpg_residuals_df, "mpg_fit.residuals", "cars_df.am_fctr"))

#print(myplot_scatter(mpg_residuals_df, "cars_df.am_fctr", "mpg_fit.residuals", 
#                     colorcol_name="cars_df.am_fctr"))
```  
The residuals of the regression does not appear biased with mean close to `0`. However, there is some minor residual heteroskedacity. Let's examine if the residuals are influenced by features not in this model.

```{r q2_3, cache=TRUE}
#print(myplot_scatter(cars_df, "am_fctr", "mpg", colorcol_name="am_fctr"))

mpg_fit_df <- data.frame(mpg.predict=predict(mpg_fit, cars_df), mpg=cars_df$mpg, 
                         am_fctr=cars_df$am_fctr, mpg.residuals=mpg_fit$residuals)
#print(summaryBy(mpg_fit.residuals ~ cars_df.am_fctr, data=mpg_residuals_df, FUN=mean))
print(myplot_violin(mpg_fit_df, "mpg.residuals", "am_fctr"))
print(myplot_scatter(mpg_fit_df, "mpg.predict", "mpg.residuals"))
mpg_fit_df <- cbind(mpg_fit_df, cars_df[, feats_fctr[feats_fctr != "am_fctr"]])
mpg_fit_df <- cbind(mpg_fit_df, cars_df[, feats_num[feats_num != "mpg"]])
print(myplot_scatter(mpg_fit_df, "mpg.predict", "mpg.residuals", colorcol_name="cyl_fctr"))
print(myplot_scatter(mpg_fit_df, "mpg.predict", "mpg.residuals", colorcol_name="vs_fctr"))
print(myplot_scatter(mpg_fit_df, "mpg.predict", "mpg.residuals", colorcol_name="gear_fctr"))
print(myplot_scatter(mpg_fit_df, "mpg.predict", "mpg.residuals", colorcol_name="carb_fctr"))

#print(myplot_scatter(mpg_fit_df, "mpg.predict", "mpg.residuals", colorcol_name="disp"))
#print(myplot_scatter(mpg_fit_df, "mpg.predict", "mpg.residuals", colorcol_name="hp"))

#print(myplot_scatter(mpg_residuals_df, "cars_df.am_fctr", "mpg_fit.residuals", 
#                     colorcol_name="cars_df.am_fctr"))
```  

Clearly, other features can provide additional insights into the relationship between mpg & transmission.
```{r q2_3_1, cache=TRUE}
mpg_cor_df <- data.frame(mpg.cor=cor(cars_df[, c(gsub("_fctr", "", feats_fctr), 
                                                 feats_num)], cars_df[, "mpg"]))
mpg_cor_df$mpg.cor.abs <- abs(mpg_cor_df$mpg.cor)
mpg_cor_df <- orderBy(~ -mpg.cor.abs, data=mpg_cor_df)
myprint_df(mpg_cor_df)
```  
Let's add features to the simple model (mpg_fit) in order of the features correlated with mpg (highest goes first)

```{r q2_4, cache=TRUE}
mpg_wt_fit <- update(mpg_fit, mpg ~ am_fctr + wt, data=cars_df)
mpg_wt_i_fit <- update(mpg_fit, mpg ~ am_fctr * wt, data=cars_df)
print(anova(mpg_fit, mpg_wt_fit, mpg_wt_i_fit))
print(summary(mpg_wt_i_fit))

mpg_wt_i_fit_df <- data.frame(mpg.predict=predict(mpg_wt_i_fit, cars_df), 
                              mpg=cars_df$mpg, 
                              am_fctr=cars_df$am_fctr, wt=cars_df$wt,
                              mpg.residuals=mpg_wt_i_fit$residuals)
print(myplot_violin(mpg_wt_i_fit_df, "mpg.residuals", "am_fctr"))
print(myplot_scatter(mpg_wt_i_fit_df, "mpg.predict", "mpg.residuals"))
plot(mpg_wt_i_fit, which=1)
plot(mpg_wt_i_fit, which=2)
plot(mpg_wt_i_fit, which=3)
plot(mpg_wt_i_fit, which=4)
plot(mpg_wt_i_fit, which=5)
plot(mpg_wt_i_fit, which=6)
```
The heteroskedacity continues to exist, but Adj-Rsq is 0.8151. Let's add cyl_fctr which is the next highest correlated feature with mpg.

```{r q2_4_F, cache=TRUE}
mpg_mean_wt_i_fit <- update(mpg_fit, mpg ~ am_fctr * I(wt - mean(wt)), data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_mean_wt_i_fit))
print(summary(mpg_mean_wt_i_fit))

mpg_wt_i_fit_df <- data.frame(mpg.predict=predict(mpg_wt_i_fit, cars_df), 
                              mpg=cars_df$mpg, 
                              am_fctr=cars_df$am_fctr, wt=cars_df$wt,
                              mpg.residuals=mpg_wt_i_fit$residuals)
print(myplot_violin(mpg_wt_i_fit_df, "mpg.residuals", "am_fctr"))
print(myplot_scatter(mpg_wt_i_fit_df, "mpg.predict", "mpg.residuals"))
plot(mpg_wt_i_fit, which=1)
plot(mpg_wt_i_fit, which=2)
plot(mpg_wt_i_fit, which=3)
plot(mpg_wt_i_fit, which=4)
plot(mpg_wt_i_fit, which=5)
plot(mpg_wt_i_fit, which=6)
```

```{r q2_5, cache=TRUE}
mpg_cyl_fit <- update(mpg_wt_i_fit, mpg ~ am_fctr + wt + cyl_fctr, data=cars_df)
mpg_cyl_i_fit <- update(mpg_wt_i_fit, mpg ~ am_fctr * wt + am_fctr * cyl_fctr, data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_cyl_fit, mpg_cyl_i_fit))
print(summary(mpg_cyl_i_fit))
```

cyl_fctr does not seem to be a valuable addition to the model. 

```{r q2_6, cache=TRUE}
mpg_disp_fit <- update(mpg_wt_i_fit, mpg ~ am_fctr + wt + disp, data=cars_df)
mpg_disp_i_fit <- update(mpg_wt_i_fit, mpg ~ am_fctr * wt + am_fctr * disp, data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_disp_fit, mpg_disp_i_fit))
print(summary(mpg_disp_i_fit))
```
Even though mpg_disp_i_fit is "better" than mpg_wt_i_fit, most of the coefficients of mpg_disp_i_fit are not significant. So, we discard these models.

```{r q2_7, cache=TRUE}
mpg_hp_fit <- update(mpg_wt_i_fit, mpg ~ am_fctr + wt + hp, data=cars_df)
mpg_hp_i_fit <- update(mpg_wt_i_fit, mpg ~ am_fctr * wt + am_fctr * hp, data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_hp_fit, mpg_hp_i_fit))
print(summary(mpg_hp_i_fit))

mpg_hp_i_fit_df <- data.frame(mpg.predict=predict(mpg_hp_i_fit, cars_df), 
                              mpg=cars_df$mpg, 
                              am_fctr=cars_df$am_fctr, wt=cars_df$wt, hp=cars_df$hp,
                              mpg.residuals=mpg_hp_i_fit$residuals)
print(myplot_violin(mpg_hp_i_fit_df, "mpg.residuals", "am_fctr"))
print(myplot_scatter(mpg_hp_i_fit_df, "mpg.predict", "mpg.residuals"))
```

```{r q2_8, cache=TRUE}
mpg_drat_fit <- update(mpg_hp_i_fit, mpg ~ am_fctr + wt + hp + drat, data=cars_df)
mpg_drat_i_fit <- update(mpg_hp_i_fit, 
                         mpg ~ am_fctr * wt + am_fctr * hp + am_fctr * drat, 
                         data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_hp_i_fit, mpg_drat_fit, mpg_drat_i_fit))
print(summary(mpg_drat_fit))
```
am_fctr coeff becomes unreliable

```{r q2_9, cache=TRUE}
mpg_vs_fit <- update(mpg_hp_i_fit, mpg ~ am_fctr + wt + hp + vs, data=cars_df)
mpg_vs_i_fit <- update(mpg_hp_i_fit, 
                         mpg ~ am_fctr * wt + am_fctr * hp + am_fctr * vs, 
                         data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_hp_i_fit, mpg_vs_fit, mpg_vs_i_fit))
print(summary(mpg_vs_fit))
```
am_fctr coeff becomes unreliable

```{r q2_10, cache=TRUE}
mpg_carb_fit <- update(mpg_hp_i_fit, mpg ~ am_fctr + wt + hp + carb_fctr, data=cars_df)
mpg_carb_i_fit <- update(mpg_hp_i_fit, 
                         mpg ~ am_fctr * wt + am_fctr * hp + am_fctr * carb_fctr, 
                         data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_hp_i_fit, mpg_carb_fit, mpg_carb_i_fit))
#print(summary(mpg_vs_fit))
```
anova fails
```{r q2_11, cache=TRUE}
mpg_gear_fit <- update(mpg_hp_i_fit, mpg ~ am_fctr + wt + hp + gear_fctr, data=cars_df)
mpg_gear_i_fit <- update(mpg_hp_i_fit, 
                         mpg ~ am_fctr * wt + am_fctr * hp + am_fctr * gear_fctr, 
                         data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_hp_i_fit, mpg_gear_fit, mpg_gear_i_fit))
print(summary(mpg_gear_i_fit))
```
Anova works; some coeffs are NA but tolerable

```{r q2_12, cache=TRUE}
mpg_qsec_fit <- update(mpg_hp_i_fit, mpg ~ am_fctr + wt + hp + qsec, data=cars_df)
mpg_qsec_i_fit <- update(mpg_hp_i_fit, 
                         mpg ~ am_fctr * wt + am_fctr * hp + am_fctr * qsec, 
                         data=cars_df)
print(anova(mpg_fit, mpg_wt_i_fit, mpg_hp_i_fit, mpg_qsec_fit, mpg_qsec_i_fit))
print(summary(mpg_qsec_i_fit))
print(summary(mpg_qsec_fit))
```
Nothing works
```{r q2_B_1, cache=TRUE}
am_cor_df <- data.frame(am.cor=cor(cars_df[, 
                                           c(gsub("_fctr", "", 
                                                  feats_fctr[feats_fctr != "am_fctr"]), 
                                                 feats_num)], cars_df[, "am"]))
am_cor_df$am.cor.abs <- abs(am_cor_df$am.cor)
am_cor_df <- orderBy(~ am.cor.abs, data=am_cor_df)
myprint_df(am_cor_df)
```

```{r q2_B_2, cache=TRUE}
mpg_carb_fit <- update(mpg_fit, mpg ~ am_fctr + carb_fctr, data=cars_df)
mpg_carb_i_fit <- update(mpg_fit, mpg ~ am_fctr * carb_fctr, data=cars_df)
print(anova(mpg_fit, mpg_carb_fit, mpg_carb_i_fit))
print(summary(mpg_carb_fit))
```

```{r q2_B_3, cache=TRUE}
mpg_gear_fit <- update(mpg_fit, mpg ~ am_fctr + gear_fctr, data=cars_df)
mpg_gear_i_fit <- update(mpg_fit, mpg ~ am_fctr * gear_fctr, data=cars_df)
print(anova(mpg_fit, mpg_gear_fit, mpg_gear_i_fit))
print(summary(mpg_gear_fit))
```

```{r}
print(anova(mpg_fit, mpg_wt_i_fit, mpg_carb_i_fit))
```

```{r print_sessionInfo, echo=FALSE}
sessionInfo()
```